Poisson approximate likelihood compared to the particle filter

### 2024-09-17 ms.Rnw
some edit

% Maximization Round 1 - Full rinit

@@

% Continue: Maximization Round 2 - Full rinit

@@

% Continue: Maximization Round 3 - Full rinit

@@

% Maximization Round 1 - Stocks rinit

@@

% Continue: Maximization Round 2 - Stocks rinit

@@

% Continue: Maximization Round 3 - Stocks rinit

@@

@@

@@

@@

\begin{table}[ht] % The placement specifier can be [h!tbp]
\centering 
\caption{AIC for the OvOv rotavirus model, computed using two filtering methods. PAL is the Poisson approximate likelihood, implemented using the code of WWR. PF is the particle filter, implemented using the R package pomp \citep{pomppackagepaper}. Lines 1, 2 and 7 are from WWR, and the remainder are our own computations. We used $5 \times 10^4$ particle for both PF and PAL. PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL due to its lower Monte~Carlo variance. PF was maximized using iterated filtering, following the approach of SBH. PAL was maximized using coordinate gradient descent, using the code of WWR.
} 
\label{tab:ovovrealdata}
\begin{tabular}{lllr} 
  \hline
  & Method & Result & AIC  \\
  \hline
  1. & PF & Table~1 of WWR, originally from SBH &  \Sexpr{myround(20134.38,tab_signif)}
  \\
  2. & PAL & Table~1 of WWR & \Sexpr{myround(13778.08,tab_signif)}
  \\
  3. & PAL & Re-evaluating the data, model and parameters from WWR &
    \Sexpr{myround(-2*palfilter_lik_list_wwr$log_lik + 2*7 ,tab_signif)} 
  \\
  4. & PF & Data, model and parameters matching 3. &
    \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)) + 2*7,tab_signif)}
  \\
  5. & PF & Initial values changed to match SBH &
     \Sexpr{myround(-2*est_mle_vec_sbh["logLik"] + 2*7,tab_signif)}
  \\
  6. & PAL & Data, model and parameters matching 5.  &
     \Sexpr{myround(-2*palfilter_mif2_mle$log_lik + 2*7 ,tab_signif)}
  \\   
  7. & ARMA & Sec.~7.3.2 of WWR & 23043
  \\
  8. & ARMA & Data from WWR &
    \Sexpr{myround(wwr_benchmark,tab_signif)} 
  \\
 \hline
\end{tabular}
\end{table}

@@

Also contained additional results reinforcing the conclusions presented in this article.
The source code for this article is available at \url{https://github.com:ionides/pal-vs-pf} and archived at Zenodo.
An extended description of our methods, together with additional numerical results, is provided by \cite{hao24}.

>>>>

% Maximization Round 1 - Full rinit: also estimate initial values

@@

% Round 2 - Full rinit

@@

% Round 3 - Full rinit

@@

% Maximization Round 1 - Fix initial values with Stocks' rinit

@@

% Round 2 - Stocks rinit

@@

% Round 3 - Stocks rinit

@@

% Round 4 - Stocks rinit

@@

% Round 5 - Stocks rinit

@@

% Round 6 - Stocks rinit

@@

\begin{table}[ht] % The placement specifier can be [h!tbp]
\centering 
\caption{AIC for the OvOv rotavirus model, computed using two filtering methods. PAL is the Poisson approximate likelihood, implemented using the code of WWR. PF is the particle filter, implemented using the R package pomp \citep{pomppackagepaper}. Lines 1, 2 and 7 are from WWR, and the remainder are our own computations. We used $5 \times 10^4$ particle for both PF and PAL. PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL due to its lower Monte~Carlo variance. PF was maximized using iterated filtering, following the approach of SBH. PAL was maximized using coordinate gradient descent, using the code of WWR.
} 
\label{tab:ovovrealdata}
\begin{tabular}{lllllr} 
  \hline
  & Method & Data & Model & Parameters                          & AIC  \\
  \hline
  1. & PF  & SBH  & SBH   & Table~1 of WWR, originally from SBH &  \Sexpr{myround(20134.38,tab_signif)}
  \\
  2. & PAL & WWR  & WWR   & Table~1 of WWR                      & \Sexpr{myround(13778.08,tab_signif)}
  \\
  3. & PAL & WWR  & WWR   & Re-evaluating the data, parameters from WWR &
    \Sexpr{myround(-2*palfilter_lik_list_wwr$log_lik + 2*7 ,tab_signif)} 
  \\
  4. & PF  & WWR & WWR    & Parameters matching 3. &
    \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)) + 2*7,tab_signif)}
  \\
  5. & PF  & WWR & WWR modified   & Initial values changed to match SBH &
     \Sexpr{myround(-2*est_mle_vec_sbh["logLik"] + 2*7,tab_signif)}
  \\
  6. & PAL & WWR & WWR modified   & Data, model and parameters matching 5.  &
     \Sexpr{myround(-2*palfilter_mif2_mle$log_lik + 2*7 ,tab_signif)}
  \\   
  7. & ARMA & SBH & ARMA benchmark    & Sec.~7.3.2 of WWR & 23043
  \\
  8. & ARMA & WWR & ARMA benchmark      & ARMA benchmark &
    \Sexpr{myround(wwr_benchmark,tab_signif)} 
  \\
 \hline
\end{tabular}
\end{table}

\begin{table}[ht] % The placement specifier can be [h!tbp]
\centering 
\caption{AIC for the OvOv rotavirus model, computed using two filtering methods. PAL is the Poisson approximate likelihood, implemented using the code of WWR. PF is the particle filter, implemented using the R package pomp \citep{pomppackagepaper}. Lines 1, 2 and 7 are from WWR, and the remainder are our own computations. We used $5 \times 10^4$ particle for both PF and PAL. PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL due to its lower Monte~Carlo variance. PF was maximized using iterated filtering, following the approach of SBH. PAL was maximized using coordinate gradient descent, using the code of WWR.}
\label{tab:ovovrealdata}
\begin{tabular}{lllllr} 
  \hline
  & Method   & Data  & Model & Parameters \& Source                    & AIC  \\  
  \hline
  1. & PF      & SBH   & SBH   & Table~1 of WWR, originally from SBH &  \Sexpr{myround(20134.38,tab_signif)} \\
  2. & PAL     & WWR   & WWR   & Table~1 of WWR                      & \Sexpr{myround(13778.08,tab_signif)} \\
  3. & PAL     & WWR   & WWR   & Re-evaluating the data, parameters from WWR & \Sexpr{myround(-2*palfilter_lik_list_wwr$log_lik + 2*7 ,tab_signif)} \\
  4. & PF      & WWR   & WWR   & Parameters matching 3. & \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)) + 2*7,tab_signif)} \\
  5. & PF      & WWR   & WWR modified   & Initial values changed to match SBH & \Sexpr{myround(-2*est_mle_vec_sbh["logLik"] + 2*7,tab_signif)} \\
  6. & PAL     & WWR   & WWR modified   & Data, model and parameters matching 5. & \Sexpr{myround(-2*palfilter_mif2_mle$log_lik + 2*7 ,tab_signif)} \\
  \hline
     & Benchmark &  &  &  \\
  \hline
  7. & ARMA    & SBH   & \multicolumn{2}{c}{\centering Sec.~7.3.2 of WWR}             & 23043 \\
  8. & ARMA    & WWR   & \multicolumn{2}{c}{\centering Re-evaluating 7. with data used by WWR}                 & \Sexpr{myround(wwr_benchmark,tab_signif)} \\
  \hline
\end{tabular}
\end{table}

@@

The source code for this article is available at \url{https://github.com:ionides/pal-vs-pf} and archived at Zenodo.
An extended description of our methods, together with additional numerical results, is provided by \cite{hao24}.
